\begin{acknowledgements}

We would like to extend our heartfelt gratitude to the distinct individuals and institutions who have guided, supported, and inspired us throughout the journey of completing this thesis.

First and foremost, we acknowledge the grace of God Almighty. Our journey was anchored in the confidence that we could overcome any difficulty through Divine presence. We are eternally grateful for the wisdom, strength, and opportunities bestowed upon us.

We extend our deepest gratitude to the distinguished faculty members who have guided and supported us throughout this research:

To our Thesis Adviser, Assoc. Prof. Dante D. Dinawanao, M.Sc., for his invaluable mentorship and direction. His guidance was instrumental in shaping the foundation of this study and navigating the complexities of our research.

To Asst. Prof. Jennifer Joyce Montemayor-Maulana, M.Sc., for opening our eyes to the wonders of Neural Networks. Her expertise, coupled with her unwavering kindness, instilled in us the confidence to tackle complex algorithms and innovative solutions. Her mentorship extended beyond academic supervision, challenging us to strive for excellence.

To Asst. Prof. Julieto Perez, M.Sc., for offering new perspectives that challenged our thinking and broadened the scope of our analysis. We are grateful for the steadfast support and encouragement he demonstrated throughout our academic journey.

To Asst. Prof. John Gieveson Iglupas, MPA, for accepting a role far beyond that of a resource person. His insights opened our minds to the intricate world of policy and governance, bridging the gap between technical simulation and real-world administrative application.

Furthermore, our colleagues and fellow researchers have been an essential component of this endeavor. We appreciate the insightful comments and intellectual exchanges that refined our study process. Their constructive criticism and collaborative spirit made a major difference in the overall quality of this thesis.

We also wish to thank the Department of Computer Science and MSU-Iligan Institute of Technology for providing us with the necessary resources, datasets, and equipment. These contributions were critical in conducting our experiments and efficiently assessing our results.

To our families and friends, who have been a continual source of encouragement, we offer our deepest thanks. We appreciate your patience, understanding, and emotional support, which helped us navigate the challenges inherent in academic research.

Finally, we acknowledge the resilience and dedication we have fostered within ourselves. This thesis is a testament to our tenacity, persistence, and the sacrifices made over long hours of work. We are grateful for the personal growth and discipline gained throughout this journey.

\end{acknowledgements}